\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-18 paper.qmd
Update PBMC descriptions

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

@@

To determine whether the UMAP representation with the parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The figures in @fig-pbmc1-sc show three well-separated clusters, indicating that the suggested UMAP representation preserves the global structure (see @fig-umap-author). However, as shown in @fig-pbmc1-sc, these clusters are close to each other in \pD{}. Also, non-linear continuity patterns and high-density patches within the clusters are observed (see @fig-pbmc1-sc). Therefore, the suggested UMAP representation (see @fig-umap-author) does not accurately preserve the local structure of the PBMC3k dataset.

@@

@@

In order to find a reasonable NLDR representation for the PBMC3k dataset, we calculated the absolute error for different numbers of non-empty bins using various NLDR techniques and different parameter settings (see @fig-pbmc-abserror). After analyzing the results, we found that tSNE with a perplexity set to $30$ had the lowest error when the number of non-empty bins was $137$. Therefore, tSNE with a perplexity of $30$, which is the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

<!--compute absolute error for different parameter choices-->
```{r}
error_df_umap <- read_csv("data/pbmc3k/error_df_umap.csv", col_names = FALSE)
names(error_df_umap) <- c("error", "mse", "bin1", "bin2", "b", "b_non_empty", 
                     "method", "n_neighbors", "min_dist")

error_df_umap <- error_df_umap |>
  mutate(type = paste0("n_neighbors: ", n_neighbors, ", min_dist: ", min_dist)) |>
  select(-c(n_neighbors, min_dist))

error_df_tsne <- read_csv("data/pbmc3k/error_df_tsne.csv", col_names = FALSE)
names(error_df_tsne) <- c("error", "mse", "bin1", "bin2", "b", "b_non_empty", 
                          "method", "perplexity")

error_df_tsne <- error_df_tsne |>
  mutate(type = paste0("perplexity: ", perplexity)) |>
  select(-perplexity)

error_df <- bind_rows(error_df_umap, error_df_tsne)

error_plot_pbmc <- ggplot(error_df, 
                              aes(x = b_non_empty, 
                                  y = log(error), 
                                  group = type, 
                                  colour = type)) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 137, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#984ea3',
                              '#ff7f00','#a6cee3','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $137$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 50%

@@

As shown in @fig-tsne-suggest, there are three well-separated clusters, although they are located close to each other. Additionally, non-linear structures can also be observed within the clusters (see @fig-model-pbmc (a)). In this manner, tSNE was able to capture the data structure for the PBMC3k dataset that UMAP failed to do.

@@

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")
benchmark_pbmc <- 0.1

>>>>

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

@@

To determine whether the UMAP representation with the (hyper-)parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The @fig-umap-author shows three well-separated clusters. However, as shown in @fig-pbmc1-sc, there is no big separation between three clusters in \pD{}. Therefore, the suggested UMAP representation (@fig-umap-author) does not accurately represent the structure(s) present in PBMC3k dataset. But, when visualizing the model-in-th-data-space some unobserved structures can be seen. Some clusters have non-linear continuity patterns and high-density patches (@fig-pbmc1-sc).

@@

XXXX I think we can remove the error figure, it's not part of the story

@@

In order to find a reasonable NLDR representation for the PBMC3k dataset, the absolute error for different numbers of non-empty bins using various NLDR techniques and different (hyper-)parameter settings (@fig-pbmc-abserror) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error when there were $124$ number of non-empty bins. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

<!--compute absolute error for different parameter choices-->

```{r}
## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:36 #sqrt(NROW(training_data_pbmc)/2)

## For umap
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
pbmc_scaled_obj_umap <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj_umap$scaled_nldr

lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_umap <- diff(lim2)/diff(lim1)

error_pbmc_umap <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = umap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_umap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "UMAP_30_min_dist_0.3")

  error_pbmc_umap <- bind_rows(error_pbmc_umap, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:36 #sqrt(NROW(training_data_pbmc)/2)

## For umap
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")
pbmc_scaled_obj_umap <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj_umap$scaled_nldr

lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_umap <- diff(lim2)/diff(lim1)

error_pbmc_umap2 <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = umap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_umap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "UMAP_5_min_dist_0.01")

  error_pbmc_umap2 <- bind_rows(error_pbmc_umap2, error_df)

## For umap

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:23 #sqrt(NROW(training_data_pbmc)/2)
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
pbmc_scaled_obj_umap <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj_umap$scaled_nldr

lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_umap <- diff(lim2)/diff(lim1)

error_pbmc_umap3 <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = umap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_umap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "UMAP_15_min_dist_0.99")

  error_pbmc_umap3 <- bind_rows(error_pbmc_umap3, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:22 #sqrt(NROW(training_data_pbmc)/2)

## For tsne
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")
pbmc_scaled_obj_tsne <- gen_scaled_data(
  data = tsne_pbmc)
tsne_pbmc_scaled <- pbmc_scaled_obj_tsne$scaled_nldr

lim1 <- pbmc_scaled_obj_tsne$lim1
lim2 <- pbmc_scaled_obj_tsne$lim2
r2_tsne <- diff(lim2)/diff(lim1)

error_pbmc_tsne <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = tsne_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_tsne,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "tSNE",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "tsne_5")

  error_pbmc_tsne <- bind_rows(error_pbmc_tsne, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:26 #sqrt(NROW(training_data_pbmc)/2)

## For tsne
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")
pbmc_scaled_obj_tsne <- gen_scaled_data(
  data = tsne_pbmc)
tsne_pbmc_scaled <- pbmc_scaled_obj_tsne$scaled_nldr

lim1 <- pbmc_scaled_obj_tsne$lim1
lim2 <- pbmc_scaled_obj_tsne$lim2
r2_tsne <- diff(lim2)/diff(lim1)

error_pbmc_tsne2 <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = tsne_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_tsne,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "tSNE",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "tsne_30")

  error_pbmc_tsne2 <- bind_rows(error_pbmc_tsne2, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:46 #sqrt(NROW(training_data_pbmc)/2)

## For phate
phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
pbmc_scaled_obj_phate <- gen_scaled_data(
  data = phate_pbmc)
phate_pbmc_scaled <- pbmc_scaled_obj_phate$scaled_nldr

lim1 <- pbmc_scaled_obj_phate$lim1
lim2 <- pbmc_scaled_obj_phate$lim2
r2_phate <- diff(lim2)/diff(lim1)

error_pbmc_phate <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = phate_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_phate,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "PHATE",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "phate_5")

  error_pbmc_phate <- bind_rows(error_pbmc_phate, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:36 #sqrt(NROW(training_data_pbmc)/2)

## For trimap
trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
pbmc_scaled_obj_trimap <- gen_scaled_data(
  data = trimap_pbmc)
trimap_pbmc_scaled <- pbmc_scaled_obj_trimap$scaled_nldr

lim1 <- pbmc_scaled_obj_trimap$lim1
lim2 <- pbmc_scaled_obj_trimap$lim2
r2_trimap <- diff(lim2)/diff(lim1)

error_pbmc_trimap <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = trimap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_trimap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "TriMAP",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "trimap_12_4_3")

  error_pbmc_trimap <- bind_rows(error_pbmc_trimap, error_df)

## To initialize number of bins along the x-axis
bin1_vec_pbmc <- 2:43 #sqrt(NROW(training_data_pbmc)/2)

## For pacmap
pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
pbmc_scaled_obj_pacmap <- gen_scaled_data(
  data = pacmap_pbmc)
pacmap_pbmc_scaled <- pbmc_scaled_obj_pacmap$scaled_nldr

lim1 <- pbmc_scaled_obj_pacmap$lim1
lim2 <- pbmc_scaled_obj_pacmap$lim2
r2_pacmap <- diff(lim2)/diff(lim1)

error_pbmc_pacmap <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_pbmc) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = pacmap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_pacmap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_pbmc,
    df_bin = df_bin_pbmc,
    training_data = training_data_pbmc,
    newdata = NULL,
    type_NLDR = "PaCMAP",
    col_start = "PC_") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_pbmc),
           method = "pacmap")

  error_pbmc_pacmap <- bind_rows(error_pbmc_pacmap, error_df)

@@

```{r}
error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)
```

```{r}
# read the png file from device 
tsne_5_img <- readPNG("figures/pbmc3k/tsne_5.png", native = TRUE) 
phate_5_img <- readPNG("figures/pbmc3k/phate_5.png", native = TRUE) 
umap_img <- readPNG("figures/pbmc3k/author_view.png", native = TRUE) 
tsne_30_img <- readPNG("figures/pbmc3k/tsne_30.png", native = TRUE)

## Add arrows
arrow_df_pbmc <- tibble(x = c(94, 245, 101, 150),
             xend = c(96, 263, 98, 180),
             y = c(10.06, 9.72, 9.88, 9.855),
             yend = c(10.14, 9.8, 9.85, 10))

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 124, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black") +
    inset_element(p = tsne_5_img, 
                  left = 1.7, 
                  bottom = 0.16, 
                  right = 0, 
                  top = 0.31) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.65) +                   
    inset_element(p = umap_img, 
                  left = 1.3, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.03, 
                  right = 0, 
                  top = 0.23) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "none",
        legend.text = element_text(size = 7))

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $124$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

@@

As shown in @fig-tsne-suggest, there are three well-separated clusters, although they are located close to each other. Additionally, non-linear structures can also be observed within the clusters (@fig-tsne-suggest). This demonstrates that tSNE accurately captures the data structure of the PBMC3k dataset, which UMAP did not achieve.

@@

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 16
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")
#benchmark_pbmc <- 0.1

### 2024-07-18 paper.tex
Changed Figure 7
Update text with constraints for hex binning
Update PBMC descriptions

The value for \(b_2\) is computed by fixing \(b_1\). Considering the
upper bound of the first NLDR component, \(a_1 > \frac{1+q}{b_1 -3}\).
Similarly, for the second NLDR component,
\(a_2 > \frac{r_2(1 + q)}{(b_2 - 2)}\). Since
\(a_2 = \frac{\sqrt(3)}{2}a_1\) for regular hexagons,
\(a_1 > \frac{2r_2(1 + q)}{\sqrt{3}(b_2 - 2)}\). This is a linear
optimization problem. Therefore, the optimal solution must occur on a
vertex. Therefore,
\(b_2 = \Big\lceil2 +\frac{2r_2(b_1 - 3)}{\sqrt{3}}\Big\rceil\).

@@

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a)
\(b = 112 \text{ } (8, \text{ }14)  \text{ }(q = 0.1)\), (b)
\(b = 220 \text{ } (11, \text{ }20) \text{ }(q = 0.075)\), and (c)
\(b = 322 \text{ } (14, \text{ }26) \text{ }(q = 0.065)\). Color
indicates standardized counts, dark indicating high count and light
indicates low count. At the smallest bin size the data segregates into
two separate groups, suggesting this is too many bins. Using the MSE of
the model fit in \(p-D\) helps decide on a useful choice of number of
bins.}

@@

\caption{\label{fig-umap-author}Reproduction of plot published in Chen
et al.~(2023) showing a \(2\text{-}D\) layout from UMAP applied for the
PBMC3k dataset. The parameter settings, beyond the defaults, are 30
nearest neighbors and minimum distance equal to 0.3. We use our
model-in-the-data-space to assess whether this is an accurate
representation of structure present in the high-dimensional data, or if
it is misleading.}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the \pD{} data.
The figures in Figure~\ref{fig-pbmc1-sc} show three well-separated
clusters, indicating that the suggested UMAP representation preserves
the global structure (see Figure~\ref{fig-umap-author}). However, as
shown in Figure~\ref{fig-pbmc1-sc}, these clusters are close to each
other in \pD{}. Also, non-linear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1-sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umap-author}) does not accurately preserve the
local structure of the PBMC3k dataset.

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, we calculated the absolute error for different numbers of
non-empty bins using various NLDR techniques and different parameter
settings (see Figure~\ref{fig-pbmc-abserror}). After analyzing the
results, we found that tSNE with a perplexity set to \(30\) had the
lowest error when the number of non-empty bins was \(137\). Therefore,
tSNE with a perplexity of \(30\), which is the default parameter
setting, is considered as a reasonable representation for the PBMC3k
dataset.

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pbmc-abserror-1.pdf}

@@

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(137\), the
lowest error occurred with the parameters perplexity: \(30\).}

@@

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although they are located close to each other.
Additionally, non-linear structures can also be observed within the
clusters (see Figure~\ref{fig-model-pbmc} (a)). In this manner, tSNE was
able to capture the data structure for the PBMC3k dataset that UMAP
failed to do.

>>>>

The value for \(b_2\) is computed by fixing \(b_1\). Considering the
upper bound of the first NLDR component, \(a_1 > \frac{1+2q}{b_1 -1}\).
Similarly, for the second NLDR component,
\(a_2 > \frac{r_2 + q(1 + r_2)}{(b_2 - 1)}\). Since
\(a_2 = \frac{\sqrt(3)}{2}a_1\) for regular hexagons,
\(a_1 > \frac{2[r_2 + q(1 + r_2)]}{\sqrt{3}(b_2 - 1)}\). This is a
linear optimization problem. Therefore, the optimal solution must occur
on a vertex. Therefore,
\(b_2 = \Big\lceil1 +\frac{2[r_2 + q(1 + r_2)](b_1 - 1)}{\sqrt{3}(1 + 2q)}\Big\rceil\).

@@

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a)
\(b = 91 \text{ } (7, \text{ }13)  \text{ }(q = 0.12)\), (b)
\(b = 220 \text{ } (11, \text{ }20) \text{ }(q = 0.07)\), and (c)
\(b = 527 \text{ } (17, \text{ }31) \text{ }(q = 0.05)\). Color
indicates standardized counts, dark indicating high count and light
indicates low count. At the smallest bin size the data segregates into
two separate groups, suggesting this is too many bins. Using the MSE of
the model fit in \(p-D\) helps decide on a useful choice of number of
bins.}

@@

\caption{\label{fig-umap-author}Reproduction of plot published in Chen
et al.~(2023) showing a \(2\text{-}D\) layout from UMAP applied for the
PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are
30 nearest neighbors and minimum distance equal to 0.3. We use our
model-in-the-data-space to assess whether this is an accurate
representation of structure present in the high-dimensional data, or if
it is misleading.}

@@

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\pD{} data. The Figure~\ref{fig-umap-author} shows three well-separated
clusters. However, as shown in Figure~\ref{fig-pbmc1-sc}, there is no
big separation between three clusters in \pD{}. Therefore, the suggested
UMAP representation (Figure~\ref{fig-umap-author}) does not accurately
represent the structure(s) present in PBMC3k dataset. But, when
visualizing the model-in-th-data-space some unobserved structures can be
seen. Some clusters have non-linear continuity patterns and high-density
patches (Figure~\ref{fig-pbmc1-sc}).

XXXX I think we can remove the error figure, it's not part of the story

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the absolute error for different numbers of non-empty bins
using various NLDR techniques and different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-abserror}) were calculated. After analyzing the
results, it was found that tSNE with default (hyper-)parameter setting
(perplexity: \(30\)) achieved the lowest error when there were \(124\)
number of non-empty bins. Therefore, tSNE with a perplexity value set to
\(30\), the default parameter setting, is considered as a reasonable
representation for the PBMC3k dataset.

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pbmc-abserror-1.pdf}

@@

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(124\), the
lowest error occurred with the parameters perplexity: \(30\).}

@@

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although they are located close to each other.
Additionally, non-linear structures can also be observed within the
clusters (Figure~\ref{fig-tsne-suggest}). This demonstrates that tSNE
accurately captures the data structure of the PBMC3k dataset, which UMAP
did not achieve.
